From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-29 Pi+/paper/finite.tex
Rework up to immersion

From grade school combinatorics, we know that there are $\fac{n}$ permutations on a finite set with $n$ elements. For
every $n$, we thus want to produce a finite set with cardinality $\fac{n}$, and a bijection with $\Aut[\Fin[n]]$. There
are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called "subexcedant sequences",
or "factoriadics", which give the factorial number system. They are a particularly convenient way of representing
permutations on a computer - since they have exactly the right cardinality, they are bitwise-optimal.

Formally, $\Lehmer[n]$ is an $n+1$-element tuple, where the position $k \leq n$ stores an element of $\Fin[k]$. Since
the 0-th position is trivial, in practice, it is ignored, and the sequence starts at
1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.\todo{There are more examples - listed
  in~\cite{duboisTestsProofsCustom2018}}. We can define $\Lehmer$ in two equivalent ways, by a simple recursion on
$\Nat$, or as a type family generated by two constructors.

@@

Suppose we have a permutation $\sigma$ on an $n$-element set $\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it
as follows. $\Lehmer[n]$ is a $n$-element tuple. Then, At position $k$, we put the number of inversions of the element
$\el{k}$ in $p$, i.e. the number of elements smaller than $\el{k}$ occurring after $\el{k}$. As an example, consider the
following tabulated presentation of the permutation:

@@

\[
  \sigma =
  \begin{array}{ccccccccccccccc}
    | & 0      & | & 1      & | & 2      & | & 3      & | & 4      & | \\
    \hline                                                             \\
    | & \el{2} & | & \el{0} & | & \el{1} & | & \el{4} & | & \el{3} & | \\
    \hline                                                             \\
  \end{array}
\]

@@

The element $\el{0}$ has 0 inversions, because there are no elements smaller than $\el{0}$ occurring after it. In fact,
there can be no elements smaller than $\el{0}$ at all, so the type at the first position of the Lehmer code tuple is
$\unit$.

The element $\el{1}$ has 0 inversions as well, since elements occurring after it in the permutation are $\el{4}$ and
$\el{2}$. There is only one different case, if $\el{1}$ appeared before $\el{0}$, it would have 1 inversion. This is why
the type of the second component of the Lehmer code is $\Fin[2]$.

The element $\el{2}$ has 2 inversions, because both $\el{0}$ and $\el{1}$ occur after it in the permutation. The element
$\el{3}$ occurs as the last one, so it has 0 inversions. The element $\el{4}$ has 1 inversion, with the element
$\el{3}$.

Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 0, 2, 0, 1)$.

To reconstruct the tabulated presentation of the permutation from the Lehmer code, we perform an algorithm similar to
\emph{insertion sort}. Starting from the left-most position of the tuple $l$, we'll read the value $v$, insert the new
element at the end of the newly created list, and shift it backward $v$ places.

@@

>>>>

There are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called "subexcedant sequences",
or "factoriadics", which give the factorial number system. They are a particularly convenient way of representing
permutations on a computer, partly because they are bitwise-optimal: for any $n : \Nat$, the type $\Lehmer[n]$ hass the cardinality $\fac{n}$, and has an easy to consttuct bijection with $\Aut[\Fin[n]]$. .

Formally, we define $\Lehmer[n]$ to be an $n+1$-element tuple, where the position $k \leq n$ stores an element of $\Fin[k]$. Since
the 0-th position is trivial, in practice, it is ignored, and the sequence starts at
1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.\todo{There are more examples - listed
  in~\cite{duboisTestsProofsCustom2018}}. We can then define $\Lehmer$ in two equivalent ways, by a simple recursion on
$\Nat$, or as a type family generated by two constructors.

@@

For a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as the number of elements smaller than it appearing after in the permutation.
\todo{typify}
\begin{definition}
    Given $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by 
    $|\{j < i : \sigma(j) > \sigma(i)\}|$.
\end{definition}

It turns out that knowing just the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following tabulated presentation of the permutation:

@@

\[
  \sigma =
  \begin{array}{ccccccccccccccc}
    | & 0      & | & 1      & | & 2      & | & 3      & | & 4      & | \\
    \hline                                                             \\
    | & \el{2} & | & \el{1} & | & \el{4} & | & \el{0} & | & \el{3} & | \\
    \hline                                                             \\
  \end{array}
\]

@@

The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of $\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it is last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$). Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 1, 2, 0, 1)$.

To construct the presentation of the permutation from the Lehmer code, written using adjacent transpositions, we perform an algorithm similar to \emph{insertion sort}. Starting from the left-most position of the tuple $l$, we'll read the value $v$, insert the new
element at the end of the newly created list, and shift it backward $v$ places.

@@

\todo{So now, finally, give an explict example of a normal form, Lehmer immersion}.
